### 2018-06-21 paper/2-background.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

To support dynamic dependencies,
\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{restarting}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel can
also track changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single build task has changed. We discuss
self-tracking further in~\S\ref{sec-tracking-aspects}.

>>>>

To support dynamic dependencies, \Excel's build
algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel{} arranges the cells into a linear sequence, called the
\emph{calc chain}.  During recalculation, \Excel{} processes
cells in the calc-chain sequence, but if computing a cell C requires the
value of a cell D that has not yet beeen computed, \Excel{}
\emph{aborts} computation of C, moves D before C in the calc chain,
and resumes recalculation starting with D.  When a calc cycle is
complete, the calc chain respects all the dynamic dependencies of the
spreadsheet.  When an input value (or formula) is changed, \Excel{}
uses the final calc chain from the \emph{previous} calc cycle as its
starting point so that, in the common case where changing an input
value does not change dependencies, there are no aborts.  Notice
that calc always succeeds regardless of the initial calc chain
(barring truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel
also tracks changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single build task has changed. We discuss
self-tracking further in~\S\ref{sec-tracking-aspects}.

### 2018-06-21 paper/5-implementations.tex
Polishing from Simon
Esp early in section 5, and Excel in section 2

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i k v = Rebuilder c i k v -> Build c k v
type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
\end{minted}
That is:
\begin{itemize}
\item A \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
  it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
\item A \hs{Rebuilder} is a function that is given a key, the key's ``old'' value,
  and a \hs{Task} that can compute the value of the key if necessary. It uses
  the persistent build state \hs{i} (carried by the state monad) to decide
  whether to re-build the value.  If doing so is unnecesary, it returns the
  old value; otherwise it runs the supplied \hs{Task} to rebuild it.  In both
  cases it updates the persistent build information \hs{i} to reflect what happened.
\end{itemize}
We write a scheduler for each column of Table~\ref{tab-build-systems}, and a
rebuilder for each row; then applying the former to the latter, in any combination,
gives a correct build system.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = return (getValue k store)
            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-3mm}
\end{figure}

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. The top-level definition
\hs{make}~\hs{=}~\hs{topological}~\hs{modTimeRebuilder} is a direct translation
of the scheduler-rebuilder decomposition in Table~\ref{tab-build-systems}. Below
we describe both components, starting with \hs{modTimeRebuilder}.

\Make stores the pair \hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTime)}
as persistent build information. It comprises the \emph{current time} \hs{now}
and the map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a \hs{key}, its
current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
\hs{key} if necessary. The rebuilder returns a modified monadic task of type
\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
access to \Make's build information and uses it to decide whether to return the
current \hs{value} as is, or rebuild the \hs{key} if it is \hs{dirty}, i.e.
if it is older than any of its dependencies\footnote{Since the \hs{task} is
applicative we can statically find its dependencies using the function
\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
(in which case \hs{modTime} has no entry for it). Note that in our model we need
to update the current time and file modification time whenever the \hs{key} is
rebuilt, which in reality is taken care of by the file system.

@@

Our \Excel's model is a composition of the \hs{restarting} scheduler and the
\hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
Persistently stored build information \hs{ExcelInfo}~\hs{k} is a triple:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- Build the key
                                 lift (gets (getInfo . fst)) >>= put -- Save new traces
                                 lift (gets (getValue k . fst))      -- Fetch the value
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask fetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

>>>>

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c is ib k v = Rebuilder c ib k v -> Build c is k v
type Rebuilder c i k v = k -> v -> Task c k v -> Task (MonadState i) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
The \hs{Rebuilder} makes uses of the persistent build information \hs{ib}, while
the scheduler might augment that with further persistent information of its own,
yielding \hs{is}.

A \hs{Rebuilder} is a function which the \hs{Scheduler} uses to rebuild
a particular key, passing to it the key, the key's ``old'' value,
and a \hs{Task} that can compute the value of the key if necessary. It uses
the persistent build state \hs{i} (carried by the state monad) to decide
whether to re-build the value.  If doing so is unnecesary, it returns the
old value; otherwise it runs the supplied \hs{Task} to rebuild it.  In both
cases it updates the persistent build information \hs{i} to reflect what happened.
So a \hs{Rebuilder} wraps a \hs{Task c k v}, which unconditionally rebulds the key,
to make a \hs{Task (Monadstate i) k v}, which rebuilds the key only if necessary, and
does the necessary book-keeping. \simon{I'm puzzled by the absence of 'c' in the return type.}

These two abstractions are the key to modularity: \emph{we can combine any scheduler
  with any rebuilder, and obtain a correct build system}.
In this section we will write a scheduler for each column of Table~\ref{tab-build-systems},
and a rebuilder for each row; then combine them to obtain the build systems in the
body of the Table.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = return (getValue k store)
            (newValue, newInfo) = runState (run newTask fetch) (getInfo store)
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-3mm}
\end{figure}

An implementation of \Make using our framework is shown in
Fig.~\ref{fig-make-implementation}. As promised, its definition
is just the application of a \hs{Scheduler}, \hs{topological},
to a \hs{Rebuilder}, \hs{modTimeRebuilder}.
We discuss each component in turn, starting with the rebuilder.

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTime}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the task's dependencies
(computed by \hs{dependencies}) are out of date.  If the key is
\hs{dirty}, we use `run task` to rebuild it, and update the state with
the new modification time of the key\footnote{The real \hs{make}
relies on the file system to track modification times, but we prefer
to make it explicit here.}; othewise we can just return the old vale.

@@

Our \Excel's model is a composition of the \hs{restarting} scheduler and the
\hs{approximateRebuilder}, as shown in Fig.~\ref{fig-excel-implementation}.
The persistent build information \hs{ExcelInfo}~\hs{k} is a triple:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, Map k [k]), Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, Map k [k]) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (dirtyKeys, deps) <- get
    let dirty = key `Set.member` dirtyKeys ||
                case Map.lookup key deps of Nothing -> True
                                            Just ks -> any (`Set.member` dirtyKeys) ks
    if not dirty then return value else do
        put (Set.insert key dirtyKeys, deps)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad i (i, Chain k) k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [target | target `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    [@@]       = return [@@]
    go done (key:ks) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) ks
        Just task -> do
            store <- get
            let newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [key]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, [@@]) . putValue key newValue
                    (key :) <$> go (Set.insert key done) ks
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key value (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState (build target) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) (@@)
    build key = case tasks key of
        Nothing -> return (@@)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do lift (build k)                      -- Build the key
                                 lift (gets (getInfo . fst)) >>= put -- Save new traces
                                 lift (gets (getValue k . fst))      -- Fetch the value
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask fetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}
